Accelerated evolution of convective simulations

### 2018-06-15 tex/bvp_paper.tex
finishes iteration of paper

\begin{abstract}
High Rayleigh number, turbulent convection is a classic system with a large timescale
separation between flow speeds and equilibration time.
In this paper, we present a method of achieving Accelerated Evolution (AE) of convective simulations which
fast-forwards through the long thermal evolution of these systems, and we test the
validity of this method. The AE procedure involves measuring the dynamics of convection
early in a simulation and using its characteristics to adjust the mean thermodynamic
profile within the domain towards its evolved state. We study \RB convection as a test case for AE.  
Evolved flow properties of AE solutions are measured to be within a few percent
of solutions which are reached through standard evolution over a full thermal timescale. 
We conclude with a discussion of useful extensions of AE, including 
stratified, compressible convection or stellar structure models.
\end{abstract}
\maketitle

@@

Unfortunately, astrophysical convective systems are stiff in more than one 
timescale. Specifically, the Peclet number (Pe), the ratio of the thermal
diffusion timescale to the convective velocity timescale, is large.
In a high Pe system, many convective timescales must pass before 
the domain reaches thermal equilibrium, a steady state in which
evolution of the thermal structure of the convective region has ceased.
As the timestep size of implicit methods is bound to the fastest nonlinear flows,
(convection) fully implicit methods cannot be used to address this form of stiffness
\cite{viallet&all2011, viallet&all2013, viallet&all2016}.

@@

In order to adjust the temperature profile to achieve AE, we calculate the total flux,
F$_{\text{tot}} =$ F$_{\text{E}}$ + F$_{\kappa}$, and then derive the profiles
\vspace{-0.5cm}
\begin{equation}
f_{\text{E}}(z) = \frac{\text{F}_{\text{E}}}{\text{F}_{\text{tot}}},\qquad
f_{\kappa}(z) = \frac{\text{F}_{\kappa}}{\text{F}_{\text{tot}}},
\label{eqn:bvp_ratios}
\end{equation}
which have the systematic asymmetries (Fig.~\ref{fig:time_trace}b) removed. These profiles describe which
parts of the atmosphere depend on convection to carry flux (where $f_{\text{E}}(z) = 1$
and $f_{\kappa}(z) = 0$).
We presume that the early convection occupies roughly the same volume as the evolved
convection, and thus that the extent of the early thermal boundary layers 
(where $f_{\kappa}(z) = 1$ and $f_{\text{E}}(z) = 0$) 
will not change significantly over the course of the atmosphere's evolution.
Under this assumption, in order to reach the converged state, 
the flux through the atmosphere must be decreased by some amount,
\vspace{-0.5cm}
\begin{equation}
\xi(z,t) \equiv \frac{\text{F}_{\text{B}}}{\text{F}_{\text{tot}}},
\label{eqn:xi}
\end{equation}
where $\text{F}_{\text{B}} = \mathcal{P}$ is the amount of flux that enters the
atmosphere at the bottom, fixed-flux boundary.
$\xi$ is a general depth-dependent scaling factor which
tells us roughly how much larger F$_\text{E}$ is in the interior than it should
be in the thermally equilibrated state (e.g., in Fig.~\ref{fig:time_trace}b,
F$_{\text{tot}} \sim$ F$_{\text{E}} \sim 15\mathcal{P}$ just below the upper boundary layer, 
but in the equilibrated state 
it should just be $\mathcal{P}$, so $\xi \sim 1/15$ at that depth).

In order to properly reduce F$_\text{E}$, we note that, upon removing 
terms which vanish due to symmetry, the convective enthalpy flux is carried
solely by the velocity field coupling to perturbations in temperature around
the mean temperature profile, or F$_\text{E} \propto \{w, T-\angles{T}_{x,y}\}$.
As the evolved enthalpy flux should be roughly $\xi$F$_\text{E}$, we reduce 
the velocity, $\bm{u}$,
and temperature perturbations about the mean state, $T-\angles{T}_{x,y}$,
 by $\sqrt{\xi}$.

To properly reduce the conductive flux, we examine the 
time-stationary state, horizontally- and time-averaged
Eqns.~(\ref{eqn:bouss_momentum}) and (\ref{eqn:bouss_energy}); after 
neglecting terms which vanish due to symmetry, these equations become
\vspace{-0.2cm}
\begin{gather}
\frac{\partial}{\partial z}\angles{\varpi}_{x,y} - \angles{T_1}_{x,y}\hat{z} = \angles{\bm{u}\times\bm{\omega}}_{x,y, \text{ ev}},
	\label{eqn:bouss_BVP_momentum}
\\
\frac{\partial}{\partial z}\text{F}_{\text{E, ev}} - \mathcal{P}\frac{\partial^2}{\partial z^2} \angles{T_1}_{x,y} = 0.
	\label{eqn:bouss_BVP_energy}
\end{gather}
Here, we construct  $\text{F}_{\text{E, ev}} = \xi \text{F}_{\text{E}}$ 
and $\angles{\bm{u}\times\bm{\omega}}_{x, y,\text{ ev}} =
\xi \angles{\bm{u}\times\bm{\omega}}_{x, y}$, where we scale 
$\angles{\bm{u}\times\bm{\omega}}_{x, y} \propto \bm{u}^2$, by $\xi$ because we
scaled $\bm{u}$ by $\sqrt{\xi}$ in order to reduce the convective enthalpy flux.
With these profiles, we solve for $\angles{\varpi}_{x,y}$ and $\angles{T_1}_{x,y}$.
Convective flows
are perturbations around a thermal profile defined by these equations in the evolved, 
statistically stationary state. Thus, the specification of
F$_{\text{conv, ev}}$ and $\angles{\bm{u}\times\bm{\omega}}_{x,y\text{, ev}}$
fully specifies the mean thermodynamic structure of the system. 
Our choice of fixing the temperature at the top boundary and the flux at the bottom
boundary ensures that there is a unique solution for $\angles{T_1}$ given the evolved
fluxes. If flux were fixed at both boundaries, there would be infinitely many
temperature solutions. If temperature were fixed at both boundaries, 
$\text{F}_{\text{B}}$ would not be precisely known \emph{a priori}.

Thus, the AE method occurs in the following steps. First, we pause a convective simulation and
construct $\xi(z,t)$ from measured flux profiles in the convective domain.
We solve a 1D boundary value problem (BVP) consisting of
Eqns.~(\ref{eqn:bouss_BVP_momentum}) \& (\ref{eqn:bouss_BVP_energy})
to obtain an evolved thermodynamic profile, and its corresponding conductive flux.
We multiply both the temperature perturbations around the mean and the
convective velocity flows by $\sqrt{\xi}$.
After adjusing the fields of a simulation in this manner, we continue timestepping forward.
For specifics on the precise implementation of the AE method, we refer
the reader to appendix \ref{appendix:recipe}.

@@

\begin{figure}[t!]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{Comparisons of the evolved thermodynamic states of an AE and SE run
at $S = 10^{5}$ are shown.  (a) Evolved horizontally- and time-averaged 
temperature profiles, as a function of height.
(b) Probability Distribution Functions (PDFs) and their integrated
Cumulative Distribution Functions (CDFs)
of point-by-point measurements of the temperature field.
(c) The percentage difference between the mean temperature profiles as a function of height.
The difference between the mean profiles is very small, O(0.5\%).
(d) $\KS{T}$, as defined in Eqn.~\ref{eqnks_profile}, is shown. The small
difference in the mean interior temperature between AE and SE
results in a large difference between the two temperature distributions near the values
of the temperature maxima.  The spread of temperature around the maxima, which includes the 
fluctuations that drive convection, are nearly identical between the two runs.
\label{fig:temp_comparison} }
\end{figure}

@@

Computational time-saving is the primary reason to use AE rather than 
evolving all solutions through SE.
In table 
\ref{table:speed}, we compare cpu-hour cost for select
2D and 3D runs in which AE and SE solutions were compared. Times reported
for AE and SE runs only include the time required to reach an equilibrated
state, and do not include the time over which measurements were taken
in the equilibrated state (e.g., the green shaded regions of
Fig.~\ref{fig:time_trace}a\&c are not included in the $S = 10^5$ times).
All simulations
were performed on Broadwell nodes on NASA's Pleiades supercomputer 
(Intel Xeon E5-2680v4 processors). The key metric which highlights the usefulness
of AE is the number of cpu-hours used for the AE run divided by cpu-hours used
for the SE run ($t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$). We see that at low
resolution and low supercriticality, $t_{\text{CPU,AE}}/t_{\text{CPU,SE}} > 1$,
and AE is not useful. However, as $S$ grows, $t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$
shrinks. At the highest supercriticalities for which AE and SE were compared
in this work, AE runs cost roughly an order of magnitude less computing time 
than SE runs.

>>>>

\begin{abstract}
High Peclet number, turbulent convection is a classic system with a large timescale
separation between flow speeds and equilibration time.
In this paper, we present a method of achieving Accelerated Evolution (AE) of convective simulations which
fast-forwards through the long thermal evolution of these systems, and we test the
validity of this method. The AE procedure involves measuring the dynamics of convection
early in a simulation and using its characteristics to adjust the mean thermodynamic
profile within the domain towards its evolved state. We study \RB convection as a test case for AE.  
Evolved flow properties of AE solutions are measured to be within a few percent
of solutions which are reached through Standard Evolution (SE) over a full thermal timescale.
At the highest values of the Rayleigh number at which we compare SE and AE,
we find that AE solutions require roughly an order of magnitude fewer cpu-hours
to compute than SE solutions.
\end{abstract}
\maketitle

@@

Unfortunately, astrophysical convective systems are stiff in more than one 
timescale. Specifically, the Peclet number (Pe), the ratio of the thermal
diffusion timescale to the convective velocity timescale, is large.
In a high Pe system, many convective timescales must pass before 
the domain reaches thermal equilibrium, a steady state in which
evolution of the thermal structure of the convective region has ceased.
As the timestep size of implicit methods is bound to the fastest nonlinear flows
(convection), fully implicit methods cannot be used to address this form of stiffness
\cite{viallet&all2011, viallet&all2013, viallet&all2016}.

@@

In order to adjust the temperature profile to achieve AE, we calculate the total flux,
F$_{\text{tot}} =$ F$_{\text{E}}$ + F$_{\kappa}$, and then derive the profiles
\vspace{-0.5cm}
\begin{equation}
f_{\text{E}}(z) = \frac{\text{F}_{\text{E}}}{\text{F}_{\text{tot}}},\qquad
f_{\kappa}(z) = \frac{\text{F}_{\kappa}}{\text{F}_{\text{tot}}},
\label{eqn:bvp_ratios}
\end{equation}
which have the systematic asymmetries (Fig.~\ref{fig:time_trace}b) removed. These profiles describe which
parts of the atmosphere depend on convection to carry flux (where $f_{\text{E}}(z) = 1$
and $f_{\kappa}(z) = 0$).
We presume that the early convection occupies roughly the same volume as the evolved
convection, and thus that the extent of the early thermal boundary layers 
(where $f_{\kappa}(z) = 1$ and $f_{\text{E}}(z) = 0$) 
will not change significantly over the course of the atmosphere's evolution.
Under this assumption, in order to reach the converged state, 
the flux through the atmosphere must be decreased by some amount,
\vspace{-0.5cm}
\begin{equation}
\xi(z,t) \equiv \frac{\text{F}_{\text{B}}}{\text{F}_{\text{tot}}},
\label{eqn:xi}
\end{equation}
where $\text{F}_{\text{B}} = \mathcal{P}$ is the amount of flux that enters the
atmosphere at the bottom, fixed-flux boundary.
For example, in Fig.~\ref{fig:time_trace}b,
F$_{\text{tot}} \approx 19\mathcal{P}$ at the upper boundary,
but in the equilibrated state 
it should just be $\mathcal{P}$, so $\xi \approx 1/19$ at that depth.

To reduce the conductive flux by $\xi$, we examine the 
time-stationary state, horizontally- and time-averaged
Eqns.~(\ref{eqn:bouss_momentum}) and (\ref{eqn:bouss_energy}); after 
neglecting terms which vanish due to symmetry, these equations become
\vspace{-0.2cm}
\begin{gather}
\frac{\partial}{\partial z}\angles{\varpi}_{x,y} - \angles{T_1}_{x,y}\hat{z} = \angles{\bm{u}\times\bm{\omega}}_{x,y, \text{ ev}},
	\label{eqn:bouss_BVP_momentum}
\\
\frac{\partial}{\partial z}\text{F}_{\text{E, ev}} - \mathcal{P}\frac{\partial^2}{\partial z^2} \angles{T_1}_{x,y} = 0.
	\label{eqn:bouss_BVP_energy}
\end{gather}
Here, we construct  $\text{F}_{\text{E, ev}} = \xi \text{F}_{\text{E}}$ 
and $\angles{\bm{u}\times\bm{\omega}}_{x, y,\text{ ev}}
= \xi\angles{\bm{u}\times\bm{\omega}}_{x, y}$.
Using these profiles and 
Eqns. (\ref{eqn:bouss_BVP_momentum}) \& (\ref{eqn:bouss_BVP_energy}),
we solve for $\angles{\varpi}_{x,y}$ and $\angles{T_1}_{x,y}$.
Our choice of fixing the temperature at the top boundary and the flux at the bottom
boundary ensures that there is a unique solution for $\angles{T_1}_{x,y}$ given the evolved
fluxes. If flux were fixed at both boundaries, there would be infinitely many
temperature solutions. If temperature were fixed at both boundaries, 
$\text{F}_{\text{B}}$ would not be precisely known \emph{a priori}.

Solving the above equations adjusts the mean thermal state and 
$\text{F}_{\kappa}$ of the atmosphere while leaving $\text{F}_{\text{E}}$
unchanged. In the  
bulk of the domain, convective enthalpy flux dominates transport, and so there we
assume that $\text{F}_{\text{tot}} \approx \text{F}_{\text{E}}$. Upon 
removing terms which vanish due to symmetry, we note that the convective
enthalpy flux is carried solely by the velocity field and perturbations in
temperature away from the mean state, $\text{F}_{\text{E}} = 
\angles{w(T - \angles{T})}$.
In order to instantaneously decrease the convective enthalpy flux in a manner
which is consistent with the conductive flux, 
we multiply both the velocity, $\bm{u}$,
and temperature perturbations about the mean state, $T-\angles{T}_{x,y}$,
 by $\sqrt{\xi}$. This scaling of the velocity field is the reason that we
multiply $\angles{\bm{u}\times\bm{\omega}}_{x, y}$, which is nondimensionally
of order $u^2/L$, by $\xi$ while constructing
$\angles{\bm{u}\times\bm{\omega}}_{x, y,\text{ ev}}$.

In general, the AE method occurs in the following steps. First, we pause a convective simulation and
construct $\xi(z,t)$ from measured flux profiles in the convective domain.
We solve a 1D boundary value problem (BVP) consisting of
Eqns.~(\ref{eqn:bouss_BVP_momentum}) \& (\ref{eqn:bouss_BVP_energy})
to obtain an evolved thermodynamic profile, and its corresponding conductive flux.
We multiply both the temperature perturbations around the mean and the
convective velocity flows by $\sqrt{\xi}$.
After adjusing the fields of a simulation in this manner, we continue timestepping forward.
For specifics on the precise implementation of the AE method, we refer
the reader to appendix \ref{appendix:recipe}.

@@

\begin{figure}[t!]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{Comparisons of the evolved thermodynamic states of an AE and SE run
at $S = 10^{5}$ are shown.  (a) Evolved horizontally- and time-averaged 
temperature profiles, as a function of height.
(b) Probability Distribution Functions (PDFs) and their integrated
Cumulative Distribution Functions (CDFs)
of point-by-point measurements of the temperature field.
(c) The percentage difference between the mean temperature profiles as a function of height.
The difference between the mean profiles is very small, O(0.5\%).
(d) $\KS{T}$, as defined in Eqn.~\ref{eqn:ks_profile}, is shown. The small
difference in the mean interior temperature between AE and SE
results in a large difference between the two temperature distributions near the values
of the temperature maxima.  The spread of temperature around the maxima, which includes the 
fluctuations that drive convection, are nearly identical between the two runs.
\label{fig:temp_comparison} }
\end{figure}

@@

Computational time-saving is the primary reason to use AE rather than 
evolving all solutions through SE.
In table 
\ref{table:speed}, we compare cpu-hour cost for select
2D and 3D runs in which both AE and SE solutions were computed. Times reported
for AE and SE runs only include the time required to reach an equilibrated
state, and do not include the time over which measurements were taken
in the equilibrated state (e.g., the green shaded regions of
Fig.~\ref{fig:time_trace}a\&c are not included in the $S = 10^5$ times).
All simulations
were performed on Broadwell nodes on NASA's Pleiades supercomputer 
(Intel Xeon E5-2680v4 processors). The key metric which highlights the usefulness
of AE is the number of cpu-hours used for the AE run divided by cpu-hours used
for the SE run ($t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$). We see that at low
resolution and low supercriticality, $t_{\text{CPU,AE}}/t_{\text{CPU,SE}} > 1$,
and AE is not useful. However, as $S$ grows, $t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$
shrinks. At the highest supercriticalities for which AE and SE were compared
in this work, AE runs cost roughly an order of magnitude less computing time 
than SE runs.
